# pre-class


Make sure you commit this often with meaningful messages. 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Standardizing a variable means subtracting the mean, and then dividing by the standard deviation. Let’s use a loop to standardize the numeric columns in the [Western Collaborative Group Study](https://clinicaltrials.gov/ct2/show/NCT00005174). This study began in 1960 with 3154 men ages 39-59, who were employed in one of 11 California based companies. They were followed until 1969 during this time, 257 of these men developed coronary heart disease (CHD). You can read this data in with the code below. You can access this dataset with the following code:

```{R}

suppressMessages(library(foreign))
wcgs <- read.dta("https://drive.google.com/uc?export=download&id=0B8CsRLdwqzbzYWxfN3ExQllBQkU")
```

The data has the following variables:



WCGS has the following variables:

-----------------------------------------------------------
Name    Description
------- -------------------------------------------
id      Subject identification number

age     Age in years

height  Height in inches

weight  Weight in lbs.

sbp     Systolic blood pressure in mm 

dbp     Diastolic blood pressure in mm Hg

chol    Fasting serum cholesterol in mm 

behpat  Behavior

  1       A1

  2       A2

  3       B3

  4       B4

ncigs   Cigarettes per day

dibpat  Behavior

1       type A

2       type B

chd69   Coronary heart disease

1       Yes

0       no

typechd Type of CHD

1       myocardial infarction or death

2       silent myocardial infarction

3       angina perctoris

time169 Time of CHD event or end of follow-up

arcus   Arcus senilis

0       absent

1       present

bmi     Body Mass Index
-----------------------------------------------------------




### Question 1: Standardize Function

A. Create a function called standardize.me() that takes a numeric vector as an argument, and returns the standardized version of the vector. 
``` {r}
library(purrr)
library(dplyr)

standarize.me = function(vec) {
  vec.sd = sd(vec)
  vec.mean = mean(vec)
  vec.final = (vec - vec.mean) / vec.sd
  return(vec.final)
}

```

B. Assign all the numeric columns of the original WCGS dataset to a new dataset called WCGS.new.
``` {r}
numeric_cols = map_lgl(wcgs, is.numeric)
WCGS.new = wcgs[,numeric_cols]
```
C. Using a loop and your new function, standardize all the variables WCGS.new dataset.
``` {r}
standardized_WCGS.new = WCGS.new
for (i in 1:ncol(WCGS.new)) {
  standardized_WCGS.new[, i] = standarize.me(WCGS.new[,i])
}
 standardized_WCGS.new
 
```

D. What should the mean and standard deviation of all your new standardized variables be? Test your prediction by running a loop
``` {r}
#Because we have standardized all of the columns, the mean for each column should be 0, and the standard deviation for each column should be 1

for(i in 1:11) {
   print(paste("The mean for column ", i, "is", round(mean(standardized_WCGS.new[,i], na.rm = T))))
   print(paste("The sd for column ", i, "is", sd(standardized_WCGS.new[,i], na.rm =  T)))
}

#as we can see, the mean's are 0 for all of the columns, and the standard deviation is 1 for all of the columns
```



### Question 2: Looping to Calculate

A. Using a loop, calculate the mean weight of the subjects separated by the type of CHD they have.
``` {r} 
mean_weight_by_CHD = data.frame(typeCHD = c("no CHD", "MI or SD", "silent MI", "angina"),
                                weight = c(0,0,0,0))
count1 = 0
count2 = 0
count3 = 0
count4 = 0


for(i in 1:nrow(wcgs)) {
  if(wcgs$typchd69[i] == "no CHD") {
    mean_weight_by_CHD[1,2] =  mean_weight_by_CHD[1,2] + wcgs$weight[i]
    count1 = count1 + 1
  } else if (wcgs$typchd69[i] == "MI or SD") {
      mean_weight_by_CHD[2,2] =  mean_weight_by_CHD[2,2] + wcgs$weight[i]
    count2 = count2 + 1
  } else if (wcgs$typchd69[i] == "silent MI") {
      mean_weight_by_CHD[3,2] =  mean_weight_by_CHD[3,2] + wcgs$weight[i]
    count3 = count3 + 1
  } else {
      mean_weight_by_CHD[4,2] =  mean_weight_by_CHD[4,2] + wcgs$weight[i]
    count4 = count4 + 1
  
  } 
}
     mean_weight_by_CHD[1,2] =  mean_weight_by_CHD[1,2]/count1
     mean_weight_by_CHD[2,2] =  mean_weight_by_CHD[2,2]/count2
     mean_weight_by_CHD[3,2] =  mean_weight_by_CHD[3,2]/count3
     mean_weight_by_CHD[4,2] =  mean_weight_by_CHD[4,2]/count4
     
mean_weight_by_CHD

#this could be so much easier with dplyr verbs!!!!




```


B. Now do the same thing, but now don’t use a loop
``` {r}
wcgs %>%
  group_by(typchd69) %>%
  summarise(mean_weight = mean(weight, na.rm = T))

#incredible
```

